Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-07-29 main.Rmd
intro changed again of choosing harmonies

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granularities. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $^{N_C} P_2$, with each of the circular granularity either mapped to facets or x-axis. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left would be very large and overwhelming for human consumption. Hence, there is a need to rank these harmonies basis how well they capture the variation in the measured variable.<!--The harmony pairs $(C_i, C_j)$ with higher average variation in the measured variable between different levels of $C_i$ and $C_j$ has a higher chance to bring out distinct characteristics of the data.<!-- Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. -->

@@

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable across different levels of the cyclic granularities. <!--across different levels of $C_j$ and integrating it for all levels of $C_i$. Distances or similarity measures have been extensively studied and applied to fields like clustering, classification and anomaly detection. ---> We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence (@Lin1991-fj) which is based on Kullback-Leibler divergence (@Kullback1951-jy). The Jensen-Shanon distance between two probability distribution $p_1$ and $p_2$ is given by $$d = [D(p_1, r) + D(p_2, r)]/2 \quad where \quad r = (p_1 + p_2)/2$$ where,
 $$D(p_1,p_2) = \int^{\infty}_{-\infty}p_1(x)log\frac{p_1(x)}{p_2(x)}\,dx$$ is the Kullback-Leibler divergence between $p_1$ and $p_2$. Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.

>>>>

\noindent The data structure suggests that we look at pairs of cyclic granularities at a time from $N_C$ cyclic granularities. Hence, all possible pairs of cyclic time granularities that could be analyzed together is $^{N_C} P_2$, with each of the circular granularity either mapped to facets or x-axis. Now, if $N_C$ is fairly large, even after excluding clashes, the list of harmonies left would be very large and overwhelming for human consumption. Hence, there is a need to rank the harmonies basis how well they capture the variation in the measured variable and additionally reduce the number of harmonies for further exploration/visualization. Displays that capture more variation within different categories in the same group would be important to bring out different patterns of the data. Hence, the idea  would be to select only those harmonies which have significant variation in the measured variable and to rank the selected harmonies, such that a harmony pair is rated more important if the variation between different levels of the x-axis variable is higher on an average across all levels of facet variables. This twofold objective is achieved by defining a statistic that essentially captures the variation in a harmony pair and a threshold below which the variation is not considered significant.

<!-- Hence, there is a need to select only those harmonies which have significant variation in the measured variable and hence might appear to be interesting when displayed. Moreover, there is a need to rank these harmonies basis how interesting they are, which could be measured through the captured amount of variation in the measured variable. -->

<!--The harmony pairs $(C_i, C_j)$ with higher average variation in the measured variable between different levels of $C_i$ and $C_j$ has a higher chance to bring out distinct characteristics of the data.<!-- Ranking could be based on how well a harmony pair displays the variation in distributions of the measured variable within and across each levels of the cyclic granularities. For each facet level, the most interesting display is the one in which the difference in the distribution of the measured variable between the x-axis categories is maximum. The harmony pair which has more variation between x-axis levels for all facet levels, on average, is likely to be a good measure of how important a harmony pair is displaying variation in the distributions. -->

@@

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable. <!--We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.-->
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence (@Lin1991-fj) which is based on Kullback-Leibler divergence (@Kullback1951-jy). The Jensen-Shanon distance between two probability distribution $p_1$ and $p_2$ is given by $$d = [D(p_1, r) + D(p_2, r)]/2 \quad where \quad r = (p_1 + p_2)/2$$ where,
 $$D(p_1,p_2) = \int^{\infty}_{-\infty}p_1(x)log\frac{p_1(x)}{p_2(x)}\,dx$$ is the Kullback-Leibler divergence between $p_1$ and $p_2$. Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.
